Report on Late Winter2013 Production: Image Differencing

### 2013-03-25 W13report.tex
Variance discussion.

We next run the SourceDetectionTask on the difference image, using a
``polarity'' of {\it both} to search for both positive and negative
deviations.  We use a nominal detection threshold of 5 sigma.  In the
case of pre--filtering, we detect on the difference image directly.
In post--filtering, we correlate the difference image with its Psf (by
design, the same as the Psf of $S$) before detection is run.  All else
being equal, the images immediately preceding the detection step
($D'$) should be {\it exactly} the same.  We expect minor differences
in the variance due to the fact that the post--filtered data will have
gone through 3 convolutions (astrometric resampling; Psf--matching to
the science image; Psf filtering for detection) while the
pre--filtered data will have gone through only two (astrometric
resampling; psf--matching to the maximum likelihood image).
Differences in the shapes and sizes of the Psf--matching kernels used
in the two cases may also cause systematic differences in the images.
Detections are merged using a grow radius of 2 pixels.

@@

The variance in the prefiltered and postfiltered images is likely to
be different due to the different types of operations that have
happened on the images.  We quantify this by taking the postfiltered
images, immediately before the detection step (i.e. $D'$), and compare
the variance plane to that coming from the prefiltered image
(Table~\ref{tab-variance2}).  For v6866601, which is a deconvolution
in the postfiltered production, we find that the variance is 70\%
higher when postfiltering.  This is consistent with the expectation
that deconvolution increases the noise properties of the image.  For
v6866602 and v6866603, we find that the variance is 3--4\% and 2--3\%
higher in the {\it prefiltered} data, respectively.  The primary
culprit for this is that the prefiltered data will have gone through 1
less convolution, meaning less of the variance will have been lost to
covariance.

To quantify the significance of these offsets, we compared the
empirical variance in the image planes with the variance tracked by
the variance planes to ascertain how much we are misestimating the
true variance.  We calculated the ratio of the interquartile range of
unmasked pixels in the image plane, multiplied by 0.741 and then
squared, with the median of unmasked pixels from the variance plane.
The results are summarized in Table~\ref{tab-variance1}.  We find that
we are marginally, but consistently, underestimating the variance in
our images by $\sim 1\%$.  This affect is smaller than the amplitudes
reported above, indicating that there are true differences in the
variances of prefiltered and postfiltered detection images.  As the
variance is higher in the prefiltered images, we will be slightly less
sensitive to true variability.

@@

\begin{table}
\centering
\begin{tabular}{clccc}
\hline
\multicolumn{4}{|c|}{Variance Ratios: Empirical to Propagated} \\
\hline
Visit    & $D_{pre} = D'_{pre}$ & $D_{post}$ & $D'_{post}$ \\
\hline
v6866601 &$1.012 \pm 0.002$&$1.018 \pm 0.005$&$0.599 \pm 0.017$ \\
v6866602 &$1.010 \pm 0.002$&$1.007 \pm 0.001$&$1.046 \pm 0.002$ \\
v6866603 &$1.015 \pm 0.003$&$1.008 \pm 0.001$&$1.045 \pm 0.003$ \\
\end{tabular}
\caption{Ratio of the empirical variance in the difference images,
  calculated via (0.741 times the interquartile range)$^2$, to the
  median of the variance plane.  In all cases the variance plane
  represents an {\it underestimate} of the true variance in the
  images.  We report the mean and RMS across all sensors for the best
  configurations.  }
\label{tab-variance1}
\end{table}

>>>>

We next run the SourceDetectionTask on the difference image, using a
``polarity'' of {\it both} to search for both positive and negative
deviations.  We use a nominal detection threshold of 5 sigma.  In the
case of pre--filtering, we detect on the difference image directly.
In post--filtering, we correlate the difference image with its Psf (by
design, the same as the Psf of $S$) before detection is run.  All else
being equal, the images immediately preceding the detection step
($D'$) should be {\it exactly} the same.  However, differences in the
shapes and sizes of the Psf--matching kernels used in the two cases
may cause systematic differences in the images.  Detections are merged
using a grow radius of 2 pixels.

@@

Because these images will have gone through multiple convolutions
before reaching the detection stage, it is expected that the
propagated per-pixel variance will not provide an exact representation
of the empirical variance in the image planes.  Because the
astrometric resampling uses a warping kernel that is designed to
preserve the noise properties of the images, we expect that the kernel
convolution and Psf--filtering will be the largest sources of error.

To quantify these errors, we first compared the empirical variance in
the image planes with the level tracked by the variance planes.  We
calculated the ratio of the interquartile range of unmasked pixels in
the image plane, multiplied by 0.741 and then squared, with the median
of unmasked pixels from the variance plane.  The former represents the
empirical variance, while the latter the propagated variance.  We
investigate their ratio (empirical/propagated) at 2 stages in the
processing: after creation of the difference image (image $D$), and
immediately preceding detection (image $D'$).  Recall that for
prefiltering, $D = D'$.  These results are summarized in
Table~\ref{tab-variance1}.  For prefiltering, we find that the
variance plane typically underestimates the true variance by $1\%$ for
all visits.  When using postfiltering, the deconvolution visit
v6866601 yields a similar underestimate in the difference image, but
for the other two visits the variance is tracked to better than 1\%,
with a small RMS.  However, when postfiltering with the Psf, the
variance becomes {\it overestimated} by nearly a factor of two for
deconvolution visit v6866601, and {\it underestimated} by $4-5\%$ for
the other visits.

We further investigate the variance properties of the difference
images immediately preceding detection, i.e. the difference image
itself in the case of prefiltering, and the difference image
correlated with its Psf in postfiltering.  These results are
summarized in Table~\ref{tab-variance2}.  We find that, empirically,
the images have very similar noise properties at the detection stage,
with the deconvolution visit having larger variance by $1\%$ but the
other visits having essentially equal noise properties, with a bias
towards the postfiltered data having slightly higher variance (while
the RMS is small, all values are greater than 1.0).  The medians of
the variance planes tell a different story though.  For the
deconvolution visit, the median per--pixel variance is $70\%$ higher.
While this will suppress the detections of false positives, it will
also significantly suppress the detection efficiency for true
variability.  For the other visits, the prefiltered per--pixel
variance is relatively {\it higher} by $3-4\%$.  This will decrease
the sensitivity of the prefiltered data to false positives; as shown
above, the rate is indeed smaller using the prefiltering pipeline.
This analysis demonstrates that this is primarily due to an {\it
  underestimate} of the per--pixel variance in the postfilter
processing.  As outlined by \cite{Price-Stacking}, this may be
accounted for by differences in pixel covariance when undergoing
multiple convolutions, which are not currently tracked.  We have not
yet definitively demonstrated that this is the case here.

@@

\begin{table}
\centering
\begin{tabular}{clccc}
\hline
\multicolumn{4}{|c|}{Variance Ratios: Empirical / Propagated} \\
\hline
Visit    & $D_{pre} = D'_{pre}$ & $D_{post}$ & $D'_{post}$ \\
\hline
v6866601 &$1.012 \pm 0.002$&$1.018 \pm 0.005$&$0.599 \pm 0.017$ \\
v6866602 &$1.010 \pm 0.002$&$1.007 \pm 0.001$&$1.046 \pm 0.002$ \\
v6866603 &$1.015 \pm 0.003$&$1.008 \pm 0.001$&$1.045 \pm 0.003$ \\
\end{tabular}
\caption{Ratio of the empirical variance in the difference images,
  calculated via (0.741 times the interquartile range)$^2$, to the
  median of the variance plane.  In all cases the variance plane
  represents an {\it underestimate} of the true variance in the
  images.  We report the mean and RMS across all sensors for the best
  configurations.  }
\label{tab-variance1}
\end{table}
